Extended Data

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
tweaks to results text, built out Fig. 3 caption

Growth milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`). 
Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and < one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).

@@

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the Spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW). CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`). The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).

@@

To test whether the negative effect of summer temperatures might offset the enhancement of growth by warmer spring temperatrues, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% (*4%? Double check this*) of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).

![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature. Chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }

>>>>

Growth milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`). 
Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).

@@

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW). CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`). The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of phenological milestones.
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 4.5 days/$^\circ$C at SCBI (p=0.001) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, at both sites, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).

@@

To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 4% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).

![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Colors indicate the correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`). Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }
